The not so widening achievement gap: an interntational comparison of the evolution of the SES achievement gap from 2000-2015

### 2017-07-29 paper/paper.Rnw
Finished some comments - only text changes

This study seeks to evaluate the evolution of the high/low SES achievement gap in the past 15 years for all PISA participant countries. This is different from previous work because it concentrates solely on 15 year old children, something not done for the achievement gap, and it attempts to capture the evolution of the achievement gap for countries with different educational systems. In addition, extend the previous findings by studying the relationship between the gap and the average performance of each country in order to detect if there's a tradeoff between efficiency and equality over time separately for each country. The advantages of this study are twofold.

@@

c) The widening/narrowing of the achievement gap has a source, which has been often studied to be related to everything from educational spending, income inequality, time allocation to students and preschool enrollment. The literature has concentrated very narrowly on whether the gap is increasing because the top performers are getting ahead, because both are distancing or because the bottom is falling behind. We shall pay particular attention to identifying the rate at which the top/bottom groups are evolving over time. This type of analysis is particularly useful in jotting down the mechanisms through which the gap is evolving. If the bottom performers are stagnated whereas the top is gaining ground, the efficiency-trade off dilemma of tracking might be much more credible than other explanations.

@@

where \begin{math}r\end{math} is the reliability score of the wave \footnote{Other proceedures multiply each country by their own reliability measure for each year-subject pair \citep{anna2016_global}. Unfortunately, PISA 2000 did not provide any reliability measure separately for each country and PISA 2015 has to yet release their own. At the moment of writing this paper, they were unavailable. For this reason we implement the analysis following the original work of \citet{reardon2011}}. Each PISA survey provides a reliability indicator which we use accordingly. This yields estimates of the true gaps, and eliminates any bias in the trend that may arise from differential reliability of the tests.

% Explain more in detail how the reliability indicator is calculated.

@@

% Replace the Y_i with the chosen y from the previos equation.
\begin{equation}
Y_i = \alpha_{j[i]} + \beta_{j[i]} * SES_i + \epsilon_i,\ \text{for} \ i \ \text{= 1, 2, ...,} \ n \ \text{within country} \ j
Y_i = \alpha_{j[i]} + \beta_{j[i]} * SES_i + \epsilon_i,\ \text{for} \ i \ \text{= 1, 2, ...,} \ n \ \text{for each country} \ j
\end{equation}

@@

We also calculate the standard error of this difference and use it to create uncertainty intervals in the estimation of time trends. We fit a multilevel rather than a linear model because by pooling the information together, we weight countries appropriately to their sample size. Given that including the SES dummy reduces the sample size considerably, we want to be able to estimate each country-difference as accurately as possible. For further exploration, we also compute the 90th/50th and 50th/10th SES gaps following the same method outlined above.
We also calculate the standard error of this difference and use it to create uncertainty intervals in the estimation of time trends. We fit a multilevel rather than an OLS model because by pooling the information together, we weight countries appropriately to their sample size. Given that including the SES dummy reduces the sample size considerably, we want to be able to estimate each country-difference as accurately as possible. For further exploration, we also compute the 90th/50th and 50th/10th SES gaps following the same method outlined above.
% Explain more in detail that it's a dummy and not three groups.

@@

\subsection{Efficiency-equality tradeoff}
<<graphing_achievement_disparity, cache = T>>=

>>>>

This study seeks to evaluate the evolution of the high/low SES achievement gap in the past 15 years for all PISA participant countries. This is different from previous work because it concentrates solely on 15 year old children, something not done for the achievement gap, and it attempts to capture the evolution of the achievement gap for countries with different educational systems. In addition, extend the previous findings by studying the relationship between the gap and the average performance of each country in order to detect if there's a tradeoff between good performance and equality over time separately for each country. The advantages of this study are twofold.

@@

c) The widening/narrowing of the achievement gap has a source, which has been often studied to be related to everything from educational spending, income inequality, time allocation to students and preschool enrollment. The literature has concentrated very narrowly on whether the gap is increasing because the top performers are getting ahead, because both are distancing or because the bottom is falling behind. We shall pay particular attention to identifying the rate at which the top/bottom groups are evolving over time. This type of analysis is particularly useful in jotting down the mechanisms through which the gap is evolving. If the bottom performers are stagnated whereas the top is gaining ground, the performance-equality trade off dilemma of tracking might be much more credible than other explanations.

@@

where \begin{math}r\end{math} is the reliability score of the wave \footnote{Other proceedures multiply each country by their own reliability measure for each year-subject pair \citep{anna2016_global}. The realibility estimates are calculated using Item Response Theory (IRT) analogues of traditional estimates of Person separation reliability such as internal consistency. Unfortunately, PISA 2000 did not provide any reliability measure separately for each country and PISA 2015 has to yet release their own. At the moment of writing this paper, they were unavailable. For this reason we implement the analysis following the original work of \citet{reardon2011}}. Each PISA survey provides a reliability indicator which we use accordingly. This yields estimates of the true gaps, and eliminates any bias in the trend that may arise from differential reliability of the tests.

@@

% Replace the Y_i with the chosen y from the previos equation.
\begin{equation}
\gamma_i = \alpha_{j[i]} + \beta_{j[i]} * SES_i + \epsilon_i,\ \text{for} \ i \ \text{= 1, 2, ...,} \ n \ \text{within country} \ j
\gamma_i = \alpha_{j[i]} + \beta_{j[i]} * SES_i + \epsilon_i,\ \text{for} \ i \ \text{= 1, 2, ...,} \ n \ \text{for each country} \ j
\end{equation}

@@

We also calculate the standard error of this difference and use it to create uncertainty intervals in the estimation of time trends. We fit a multilevel rather than a linear model because by pooling the information together, we weight countries appropriately to their sample size. Given that including the SES \emph{dummy} reduces the sample size considerably, we want to be able to estimate each country-difference as accurately as possible. For further exploration, we also compute the 90th/50th and 50th/10th SES gaps following the same method outlined above.
% Explain more in detail that it's a dummy and not three groups.

@@

\subsection{Performance-equality tradeoff}
<<graphing_achievement_disparity, cache = T>>=
